---
title: "ENVS-193DS_homework-03"
author: "Henderson Vo"
date: "2026-03-02"
format: 
  html:
    toc: TRUE
execute:
  message: FALSE
  warning: FALSE
---
[Github repository](https://github.com/hendersonvo/ENVS-193DS_homework-03#)

# Part 1. Set up tasks
```{r}
# load libraries
library(tidyverse)
library(here)
library(janitor)
library(readxl)

# read data
salinity <- read_csv(here("data", "salinity-pickleweed.csv"))
salinity_clean <- salinity |> 
  clean_names() |> 
  rename(salinity_ms_cm = salinity_m_s_cm)
personal <- read_csv(here("data", "envs193ds-personal-data.csv")) |> 
  clean_names()
```

# Part 2. Problems

## Problem 1. Slough soil salinity

### a. An appropriate test

There are two appropriate ways to test the strength of the relationship between salinity and California pickleweed biomass: Pearson's correlation test and a Spearman rank correlation test. The Pearson's correlation test is a parametric test (uses precise data values, not ranks), and it might be appropriate because our variables are continuous and we can check for normality. On the other hand, the Spearman rank correlation is a nonparametric test (uses ranks, not precise data values), and it might be appropriate because we can check for a monotonic relationship between our variables. 

### b. Create a visualization

```{r}
ggplot(data = salinity_clean, # use cleaned salinity data
       aes(x = salinity_ms_cm, # set salinity to x-axis 
           y = pickleweed)) +  # set biomass to y-axis
  geom_point(color = "turquoise3") + # change point colors
  labs(x = "Salinity (in mS/cm)", # add labels to axes and add title
       y = "California pickleweed biomass (in g)",
       title = "Scatterplot of pickleweed biomass vs. salinity") +
  theme_minimal() # change theme
```

### c. Check your assumptions and run your test

#### Checking assumptions

We want to run a Pearson's correlation test, so we will check for continuity and normality in our variables.

We already know that salinity and biomass are both continuous variables because they can take on any values in a given range. Our first assumption is correct!

Now we will check for normality.

```{r}
#| label: qq plot for salinity
ggplot(data = salinity_clean, # used cleaned salinity data
       aes(sample = salinity_ms_cm)) + # set salinity measurements against theoretical values
  geom_qq_line(color = "red") + # add a reference line, make it red
  geom_qq() + # add the rest of the qq plot
  labs(title = "QQ plot of salinity variable data") + # add a title 
  theme_minimal() # change the theme
```

Our salinity data appears to be relatively normal based on how our points (observations) roughly follow the red line (perfect normal distribution). Let's check our pickleweed biomass next.

```{r}
#| label: qq plot for biomass
ggplot(data = salinity_clean, # use cleaned salinity data
       aes(sample = pickleweed)) + # set biomass data against theoretical values
  geom_qq_line(color = "red") + # add a reference line, make it red
  geom_qq() + # add the rest of the qq plot
  labs(title = "QQ plot of pickleweed biomass variable data") + # add a title
  theme_minimal() # change the theme
```

Our pickleweed biomass observations appear to also be relatively normal because these points similarly appear to roughly follow the line.

Since our two variables both appear to be relatively normal, and they are both continuous, we can continue with the test!

#### Running the test

```{r}
cor.test(x = salinity_clean$salinity_ms_cm, # first variable
         y = salinity_clean$pickleweed, # second variable
         method = "pearson") # select type of correlation test, Pearson in this case
```

### d. Results communication

To evaluate the strength of the relationship between pickleweed biomass and soil salinity, we used a Pearson's correlation test which is designed to test the correlation between two continuous, normally distributed variables. We can see that we have found a moderate positive relationship between soil salinity and pickleweed biomass (Pearson's r = 0.53, t(21) = 2.90, p = 0.009, $\alpha$ = 0.05).

### e. Test implications

We have found that the relationship between pickleweed biomass and soil salinity is positive and moderately strong. This means that, on average, where we observe higher salinity soils, we might also expect to see higher pickleweed biomass. This implies that if restoration teams treat higher pickleweed biomass as a measure of restoration success, then they should consider having higher salinity soil at their restoration sites. 

### f. Double check your own work

```{r}
cor.test(x = salinity_clean$salinity_ms_cm, # first variable
         y = salinity_clean$pickleweed, # second variable
         method = "spearman") # select type of correlation test, Spearman in this case
```

These results also reflect a moderate positive relationship between soil salinity and pickleweed biomass (Spearman $\rho$ = 0.59, S = 824, p = 0.003, $\alpha$ = 0.05).

## Problem 2. Personal data

### a. Updating your visualizations

```{r}
ggplot(data = personal, # use personal data
       aes(x = weekday, # set x-axis to predictor variable (weekday)
           y = steps_taken, # set y-axis to response variable (steps)
           color = weekday)) + # differentiate by weekday
  geom_boxplot(width = 0.4) + # add a boxplot, adjust width
  geom_jitter(width = 0.1, # add jitter points, adjust width
              height = 0, # no height jitter
              alpha = 0.6) + # adjust transparency
  scale_color_manual(values = c("Yes" = "blue",
                                "No" = "orange")) +
  labs(x = "Weekday (Yes/No)", # add labels to axes, and add a title
       y = "Number of steps taken",
       title = "Boxplots showing the distribution of steps taken on \nweekdays vs. weekends") +
  theme_minimal() + # change the theme
  theme(legend.position = "none") # remove the legend
```


```{r}
ggplot(data = personal, # use personal data
       aes(x = weather, # set x-axis to predictor variable (weather)
           y = steps_taken,
           color = weekday)) + # differentiate by weekday
  geom_boxplot(width = 0.4) + # add a boxplot, adjust width
  geom_jitter(width = 0.1, # add jitter points, adjust width
              height = 0, # no height jitter
              alpha = 0.6) + # adjust transparency 
  facet_wrap(~ weekday) +
  scale_color_manual(values = c("Yes" = "blue",
                                "No" = "orange")) +
  labs(x = "Weather", # add labels to axes, and add a title
       y = "Number of steps taken",
       title = "Boxplots showing the distribution of steps taken on \nweekdays vs. weather types") +
  theme_minimal() + # change the theme
  theme(legend.position = "none") # remove the legend
```

```{r}
ggplot(data = personal, # use personal data
       aes(x = caffeine_comsumed_mg, # set x- and y-axes to appropriate variables
           y = steps_taken,
           color = weekday)) + # differentiate by weekday
  geom_point(alpha = 0.4) + # add points
    scale_color_manual(values = c("Yes" = "blue",
                                "No" = "orange")) +
  labs(x = "Caffeine consumed (mg)", # add labels to axes, add a title, adjust legend title
       y = "Steps taken",
       title = "Scatterplot showing steps taken vs. caffeine consumed",
       color = "Weekday") +
  theme_minimal() # change the theme
```

### b. Captions


## Problem 3. Affective visualization

## Problem 4. Statistical critique