---
title: "ENVS-193DS_homework-03"
author: "Henderson Vo"
date: "2026-03-02"
format: 
  html:
    toc: TRUE
execute:
  message: FALSE
  warning: FALSE
---
[Github repository](https://github.com/hendersonvo/ENVS-193DS_homework-03#)

# Part 1. Set up tasks
```{r}
# load libraries
library(tidyverse)
library(here)
library(janitor)
library(readxl)

# read data
salinity <- read_csv(here("data", "salinity-pickleweed.csv"))
salinity_clean <- salinity |> 
  clean_names() |> 
  rename(salinity_ms_cm = salinity_m_s_cm)
personal <- read_csv(here("data", "envs193ds-personal-data.csv"))
```

# Part 2. Problems

## Problem 1. Slough soil salinity

### a. An appropriate test

There are two appropriate ways to test the strength of the relationship between salinity and California pickleweed biomass: Pearson's correlation test and a Spearman rank correlation test. The Pearson's correlation test is a parametric test (uses precise data values, not ranks), and it might be appropriate because our variables are continuous and we can check for normality. On the other hand, the Spearman rank correlation is a nonparametric test (uses ranks, not precise data values), and it might be appropriate because we can check for a monotonic relationship between our variables. 

### b. Create a visualization

```{r}
ggplot(data = salinity_clean, # use cleaned salinity data
       aes(x = salinity_ms_cm, # set salinity to x-axis 
           y = pickleweed)) +  # set biomass to y-axis
  geom_point(color = "turquoise3") + # change point colors
  labs(x = "Salinity (in mS/cm)", # add labels to axes and add title
       y = "California pickleweed biomass (in g)",
       title = "Scatterplot of pickleweed biomass vs. salinity") +
  theme_minimal() # change theme
```

### c. Check your assumptions and run your test

#### Checking assumptions

We want to run a Pearson's correlation test, so we will check for continuity and normality in our variables.

We already know that salinity and biomass are both continuous variables because they can take on any values in a given range. Our first assumption is correct!

Now we will check for normality.

```{r}
#| label: qq plot for salinity
ggplot(data = salinity_clean, # used cleaned salinity data
       aes(sample = salinity_ms_cm)) + # set salinity measurements against theoretical values
  geom_qq_line(color = "red") + # add a reference line, make it red
  geom_qq() + # add the rest of the qq plot
  labs(title = "QQ plot of salinity variable data") + # add a title 
  theme_minimal() # change the theme
```

Our salinity data appears to be relatively normal based on how our points (observations) roughly follow the red line (perfect normal distribution). Let's check our pickleweed biomass next.

```{r}
#| label: qq plot for biomass
ggplot(data = salinity_clean, # use cleaned salinity data
       aes(sample = pickleweed)) + # set biomass data against theoretical values
  geom_qq_line(color = "red") + # add a reference line, make it red
  geom_qq() + # add the rest of the qq plot
  labs(title = "QQ plot of pickleweed biomass variable data") + # add a title
  theme_minimal() # change the theme
```

Our pickleweed biomass observations appear to also be relatively normal because these points similarly appear to roughly follow the line.

Since our two variables both appear to be relatively normal, and they are both continuous, we can continue with the test!

#### Running the test


## Problem 2. Personal data

## Problem 3. Affective visualization

## Problem 4. Statistical critique